---
title: "RNASeq"
output:
  html_document:
    theme: lumen        
    highlight: tango     
    toc: true            
    toc_float: true
    code_folding: hide 
    number_sections: true     
    df_print: paged      
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  results = 'hide',  
  message = FALSE,   
  warning = FALSE)
```

Environmental Setup

```{r library imports}
library(tidyverse)
library(DESeq2)
library(EnsDb.Hsapiens.v86)
library(EnhancedVolcano)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(kableExtra)
library(apeglm)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot)
```

Data import and EDA

```{r upload and inspect metadata}
rse <- readRDS("EwS.rds")
cd <- as.data.frame(colData(rse))
View(cd)
```

```{r}
colData(rse) <- colData(rse)[, c("sample", "run", "condition")]    # select data for analysis
rse$condition <- factor(rse$condition)                             # change raw text to factor format
rse$condition <- relevel(rse$condition, ref = "shCTR")             # assign the control factor explicitly
```

```{r}
all(rownames(colData(rse)) == colnames(assay(rse)))                # check if the metadata match count matrix
```
```{r}
assay(rse)[1:5, 1:5]    # check the genes designation
```

```{r}
# make the gene names readable for annotation
rownames(rse) <- gsub(pattern = "\\..+", replacement = "", x = rownames(rse))    
assay(rse)[1:5, 1:5]
```

```{r}
all(assay(rse) == floor(assay(rse)))    # possible to inspect element or check via function
```
### Principal Component Analysis

```{r}
dds <- DESeqDataSet(rse, design = ~ condition)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

```{r}
vsd <- vst(dds, blind = FALSE)
```

```{r}
pcaData <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(x = PC1, y = PC2, color = condition, label = name)) +
  geom_point(size = 3) + 
  xlab(paste0("PC1 - ", percentVar[1], "%")) +
  ylab(paste0("PC2 - ", percentVar[2], "%")) +
  theme_classic() + 
  ggtitle("Ewing Sarcoma PCA")
```
Principal Component Analysis (PCA) of the variance-stabilized counts shows a clear separation between control (shCTR) and knockdown (shEF1) samples along the first principal component (PC1). PC1 and PC2 together account for 99% of the total variance, confirming that the experimental condition is the primary driver of transcriptional changes. Notably, the knockdown group exhibits a secondary split along PC2, dividing the four replicates into two distinct sub-groups.

## MA Plot

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds, contrast = c("condition", "shEF1", "shCTR"))
```

```{r}
resNorm <- lfcShrink(dds = dds, res = res, type = "apeglm", coef = 2)
```

```{r}
plotMA(resNorm)
```
The MA-plot illustrates the relationship between the mean of normalized counts and the log2 fold change. Differentially expressed genes (colored in blue) are distributed across the entire range of expression levels, with a significant concentration of high-confidence candidates (indicated by triangles), exceeding a log2 fold change of ±10.The assymetry also suggests the predominant upregulation of genes.

## DEGs Table
```{r}
resdf <- as.data.frame(resNorm)
```

```{r annotation}
ens2sym <- AnnotationDbi::select(EnsDb.Hsapiens.v86, keys = keys(EnsDb.Hsapiens.v86),
                                 columns = c("SYMBOL"))
View(ens2sym)
```

```{r}
resdf <- resdf |> 
  rownames_to_column() |>  
  mutate(GENEID = gsub(rowname, pattern = "\\..+", replacement = "")) |> 
  dplyr::select(-rowname) |>  
  inner_join(y = ens2sym, by = "GENEID")
View(resdf)
```


```{r}
top_degs <- resdf |> 
  arrange(padj) |>         
  head(20) |>              
  dplyr::select(SYMBOL, GENEID, baseMean, log2FoldChange, padj) 
kable(top_degs, caption = "Top 20 Differentially Expressed Genes")
```

```{r}
EnhancedVolcano(resdf, lab = resdf$SYMBOL, pCutoff = 1e-100,
                FCcutoff = 3,
                x = "log2FoldChange", y = "padj")
```
## Top 10 over- and 10 underexpressed genes

```{r, fig.height=7, fig.width=7}
# plot heatmap of top 20 DE genes
ordered_df <- resdf[order(resdf$log2FoldChange, decreasing = TRUE), ]

# choose top10 underexpressed and top10 overexpressed
top_20_degs <- rbind(head(ordered_df, 10), tail(ordered_df, 10))
mat_top <- assay(vsd)[top_20_degs$GENEID, ]
rownames(mat_top) <- top_20_degs$SYMBOL
pheatmap(mat_top, 
         scale = "row", 
         clustering_distance_rows = "correlation",
         annotation_col = as.data.frame(colData(vsd)[, "condition", drop = FALSE]),
         main = "Top 10 Over- and Under-expressed DEGs",
         cellwidth = 30, 
         cellheight = 15)
```
## GSEA

```{r ORA and enrichment}
resdf2 <- resdf |> 
  arrange(padj) |> 
  mutate(gsea_metric = -log10(padj) * sign(log2FoldChange))

# Deal with inf
resdf2 <- resdf2 |> 
  mutate(padj = case_when(padj == 0 ~ .Machine$double.xmin,
                          TRUE ~ padj)) |> 
  mutate(gsea_metric = -log10(padj) * sign(log2FoldChange))

# Remove NAs and order by GSEA
resdf2 <- resdf2 |> 
  dplyr::filter(! is.na(gsea_metric)) |> 
  arrange(desc(gsea_metric))

# GSEA value histogram
hist(resdf2$gsea_metric, breaks = 100)
```
```{r GSEA }
# Get the ranked GSEA vector
ranks <- resdf2 |> 
  dplyr::select(SYMBOL, gsea_metric) |> 
  distinct(SYMBOL, .keep_all = TRUE) |> 
  deframe()

# Run GSEA
gene_sets <- msigdbr(
  species = "Homo sapiens", 
  category = "C2", 
  subcollection = "CP:KEGG_LEGACY"
  ) |> 
  dplyr::select(gs_name, gene_symbol)
gseares <- GSEA(geneList = ranks, 
                TERM2GENE = gene_sets,
                pvalueCutoff = 1,
                verbose = FALSE)
gsearesdf <- as.data.frame(gseares)
# Get the 'top' table
gsearesdf|> 
  arrange(desc(NES)) |> 
  slice(c(1:5, (n()-4):n())) |> 
  dplyr::select(ID, setSize, NES, p.adjust) 

# Add description
gsea_top_table$Direction <- ifelse(gsea_top_table$NES > 0, "Activated", "Suppressed")

kable(gsea_top_table, 
      row.names = FALSE,
      caption = "Top enriched KEGG pathways by effect size (NES)",
      col.names = c("Pathway ID", "Size", "NES", "Adj. P-value", "Knockdown effect"),
      digits = c(0, 0, 2, 4, 0)) 

```
## Significantly enriched Biological Processes by gene ratio
```{r GSEA dot plot}
dotplot(gseares)
```
The DSEA analysis results indicate that FL1-knockdown resulted in suppression of genes related to DNA-replication, RNA degradation, spliceosome and ribosome, which is probably a good sign and protects the cell from out of control division. On the other hand the processes involved in homeostasis, signaling (Jak-STAT signaling pathway) or in differentiation, proliferation and signalling (ECM Receptor interaction, focal adhesion), or cellular stages (hematopoietic cell lineage) are activated. 
Although the adjusted p-values for some pathways are relatively higher, their enrichment remains biologically consistent with the overall suppression of the oncogenic signature. The coordinated activation of cell adhesion and signaling pathways, coupled with the robust downregulation of the translational and replicative machinery clearly points toward the loss of the Ewing Sarcoma proliferative phenotype.
