---
title: "Caki-1 RNA Seq"
self_contained: true
output:
  html_document:
    theme: lumen        
    highlight: tango     
    toc: true            
    toc_float: true
    code_folding: hide 
    number_sections: false     
    df_print: paged   
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  fig.align = "center",
  out.width = "100%"
)
```

### Introduction

```{r upload and inspect data, results='hide'}
# Library imports
library(tidyverse)
library(DESeq2)
library(EnsDb.Hsapiens.v86)
library(EnhancedVolcano)
library(dplyr)
library(ggplot2)
library(ggplotify)
library(pheatmap)
library(apeglm)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot)
library(DT)
library(patchwork)
```

```{r upload and inspect data, results='hide'}
# Read and inspect the dataset
load("rse_gene.Rdata")
rse <- rse_gene
colnames(colData(rse_gene))
head(colData(rse)[, c("run", "title")])
```

```{r}
mapping <- data.frame(
  run = c("SRR2970559", 
          "SRR2970560", 
          "SRR2970562", 
          "SRR2970566", 
          "SRR2970569", 
          "SRR2970570", 
          "SRR2970573", 
          "SRR2970575", 
          "SRR2970578"),
  cond = c(rep("Untreated", 3), rep("CoCl2", 3), rep("Hypoxia", 3))
)
rse$condition <- factor(rep(c("Untreated", "CoCl2", "Hypoxia"), each = 3), 
                        levels = c("Untreated", "Hypoxia", "CoCl2"))
as.data.frame(colData(rse)[, c("run", "condition")])
```

```{r upload and inspect data, results='hide'}
assay(rse)[1:5, 1:5]    # check the genes designation
# make the gene names readable for annotation
rownames(rse) <- gsub(pattern = "\\..+", replacement = "", x = rownames(rse))
assay(rse)[1:5, 1:5]
all(assay(rse) == floor(assay(rse)))    # check if counts are integers
all(rownames(colData(rse)) == colnames(assay(rse)))    # check if the metadata match count matrix
colData(rse) <- colData(rse)[, c("sample", "run", "condition")] # select data for analysis
rse$condition <- relevel(rse$condition, ref = "Untreated") # assign the control factor explicitly
```

### QC and Normalization

#### PCA Plot

```{r normalization&PCA}
# Normalization
dds <- DESeqDataSet(rse, design = ~condition)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
vsd <- vst(dds, blind = FALSE)

# PCA
pcaData <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(x = PC1, y = PC2, color = condition, label = name)) +
  geom_point(size = 5) +
  xlab(paste0("PC1 - ", percentVar[1], "%")) +
  ylab(paste0("PC2 - ", percentVar[2], "%")) +
  theme_classic(base_size = 18) +
  ggtitle("Caki-1") +
  theme(plot.title = element_text(size = 22, face = "bold"))
```


#### MA Plot

```{r DESeq and MA plot}
dds <- DESeq(dds)
resultsNames(dds)
res_real_hypoxia <- results(dds, contrast = c("condition", "Hypoxia", "Untreated"))
resNorm_real_hypoxia <- lfcShrink(dds = dds, 
                     coef = "condition_Hypoxia_vs_Untreated", 
                     type = "apeglm")

res_chemical_hypoxia <- results(dds, contrast = c("condition", "CoCl2", "Untreated"))
resNorm_chemical_hypoxia <- lfcShrink(dds = dds, 
                     coef = "condition_CoCl2_vs_Untreated", 
                     type = "apeglm")
```


```{r DESeq and MA plot, fig.height=7, fig.width=10}
par(mfrow = c(1, 2))
common_ylim <- c(-2.5, 2.5)
plotMA(resNorm_real_hypoxia, main = "Real Hypoxia", ylim = common_ylim)
plotMA(resNorm_chemical_hypoxia, main = "Chemical Hypoxia (CoCl2)", ylim = common_ylim)
par(mfrow = c(1, 1))
```
The two plots are nearly identical with just a few exceptions: in CoCl2 treated samples the number of upregulated genes is even higher and they have higher LFC value, however it does not indicate that this treatment mimics the real hypoxia conditions.

### Differentially expressed genes

#### Volcano plot

```{r volcano plot}
# Global settings
p_adj_cut <- 0.01
lfc_cut <- 1
base_mean_cut <- 20

# Annotation
ens2sym <- AnnotationDbi::select(EnsDb.Hsapiens.v86,
  keys = keys(EnsDb.Hsapiens.v86),
  columns = c("SYMBOL")
)

resdf_real_hypoxia <- as.data.frame(resNorm_real_hypoxia) |>
  rownames_to_column() |>
  mutate(GENEID = gsub(rowname, pattern = "\\..+", replacement = "")) |>
  dplyr::select(-rowname) |>
  inner_join(y = ens2sym, by = "GENEID")

resdf_chemical_hypoxia <- as.data.frame(resNorm_chemical_hypoxia) |>
  rownames_to_column() |>
  mutate(GENEID = gsub(rowname, pattern = "\\..+", replacement = "")) |>
  dplyr::select(-rowname) |>
  inner_join(y = ens2sym, by = "GENEID")
```


```{r volcano plot, fig.height=7, fig.width=10}

# Use legend and subtitle from only one plot
# Real Hypoxia
p1 <- EnhancedVolcano(resdf_real_hypoxia,
  lab = resdf_real_hypoxia$SYMBOL,
  x = "log2FoldChange", y = "padj",
  pCutoff = p_adj_cut, FCcutoff = 2,
  title = "Real Hypoxia",
  subtitle = NULL, 
  caption = NULL,  
  gridlines.major = FALSE, gridlines.minor = FALSE,
  border = "full",
  legendPosition = "none" 
)

# Chemical Hypoxia
p2 <- EnhancedVolcano(resdf_chemical_hypoxia,
  lab = resdf_chemical_hypoxia$SYMBOL,
  x = "log2FoldChange", y = "padj",
  pCutoff = p_adj_cut, FCcutoff = 2,
  title = "Chemical Hypoxia (CoCl2)",
  subtitle = NULL,
  gridlines.major = FALSE, gridlines.minor = FALSE,
  border = "full"
)

# Join with one legend
(p1 | p2) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```
It is now even more obvious that chemical and real hypoxia are not the same as the DEG names are different and the volcano profiles are also different, but still both effect gene expression levels.
#### Top 20 Significant DEGs

```{r}
sig_res_real <- resdf_real_hypoxia |>
  filter(padj < p_adj_cut & abs(log2FoldChange) >= lfc_cut & baseMean >= base_mean_cut)

sig_res_chemical <- resdf_chemical_hypoxia |>
  filter(padj < p_adj_cut & abs(log2FoldChange) >= lfc_cut & baseMean >= base_mean_cut)

# find top DEGs for both conditions
top_real_names <- sig_res_real |> arrange(padj) |> head(20) |> pull(SYMBOL)
top_chem_names <- sig_res_chemical |> arrange(padj) |> head(20) |> pull(SYMBOL)

# find common unique top DEGs
all_top_genes <- unique(c(top_real_names, top_chem_names))

# join the top genes into common table to compare
comparison_table <- resdf_real_hypoxia |>
  filter(SYMBOL %in% all_top_genes) |>
  dplyr::select(SYMBOL, LFC_Real = log2FoldChange, padj_Real = padj) |>
  inner_join(
    resdf_chemical_hypoxia |> dplyr::select(SYMBOL, LFC_Chem = log2FoldChange, padj_Chem = padj),
    by = "SYMBOL"
  )

# display one table
datatable(comparison_table)
```

Table contains 37 genes, which means that these conditions only share 3 genes as top DEGs.
The LFC for DEGs also differ.

```{r heatmap, fig.height=7, fig.width=7}
# Choose the significant genes by major contrast (real hypoxia)
sig_res <- resdf_real_hypoxia |> 
  filter(padj < p_adj_cut & baseMean >= base_mean_cut)

sig_ordered <- sig_res |> arrange(padj)
# Selecting the top 10 over- and under-expressed DEGs prioritized by statistical significance 
#(lowest padj) to ensure robust expression patterns across replicates.
top_10_over <- sig_ordered |> 
  filter(log2FoldChange > 0) |> 
  head(10)

top_10_under <- sig_ordered |> 
  filter(log2FoldChange < 0) |> 
  head(10)

# Combine over- and underexpressed genes
hm_genes_id <- c(top_10_over$GENEID, top_10_under$GENEID)
hm_genes_symbols <- c(top_10_over$SYMBOL, top_10_under$SYMBOL)

# Prepare the matrix
mat_final <- assay(vsd)[hm_genes_id, ]
rownames(mat_final) <- hm_genes_symbols


pheatmap(mat_final,
         scale = "row",
         cluster_cols = FALSE, 
         annotation_col = as.data.frame(colData(vsd)[, "condition", drop = FALSE]),
         main = "Top 10 Over- and 10 Under-expressed DEGs",
         cellwidth = 30,
         cellheight = 12
)
```
Heatmap reveals two primary gene blocks: a top cluster of genes downregulated in physical hypoxia and a bottom cluster of genes upregulated across both treatment groups.
While CoCl2 effectively induces common hypoxia markers like HMOX1 and DDIT4, it fails to replicate the gene repression (blue areas) observed in the physical hypoxia samples. Stress-related genes (HSPA1A/B) show significantly higher expression levels (darker red) in the CoCl2 group compared to the physical hypoxia group.

### GSEA: Significantly enriched pathways

#### by NES (table)

#### by gene ratio (dotplot)

```{r GSEA dot plot}

# Завантажуємо KEGG
kegg_sets <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY") |>
  dplyr::select(gs_name, gene_symbol)

# Ранжуємо гени за 'stat'
ranks_real <- res_gsea_real |>
  distinct(SYMBOL, .keep_all = TRUE) |>
  arrange(desc(stat)) |>
  dplyr::select(SYMBOL, stat) |>
  deframe()

ranks_chem <- res_gsea_chem |>
  distinct(SYMBOL, .keep_all = TRUE) |>
  arrange(desc(stat)) |>
  dplyr::select(SYMBOL, stat) |>
  deframe()

# Запускаємо GSEA
gsea_real <- GSEA(ranks_real, TERM2GENE = kegg_sets, pvalueCutoff = 1, verbose = FALSE)
gsea_chem <- GSEA(ranks_chem, TERM2GENE = kegg_sets, pvalueCutoff = 1, verbose = FALSE)

# Топ-5 за padj
gsea_real_df <- as.data.frame(gsea_real) |> 
  arrange(p.adjust) |> 
  head(5)

gsea_chem_df <- as.data.frame(gsea_chem) |> 
  arrange(p.adjust) |> 
  head(5)

gsea_real_df
gsea_chem_df
```

```{r, fig.height = 10, fig.width = 17}
# Малюємо графіки
dotplot(gsea_real, showCategory = 5) + ggtitle("Top 5 KEGG Pathways: Real Hypoxia")+
dotplot(gsea_chem, showCategory = 5) + ggtitle("Top 5 KEGG Pathways: CoCl2")
```

```{r}

# Об'єднуємо результати для порівняння
gsea_real_df <- as.data.frame(gsea_real) |> mutate(Condition = "Real Hypoxia")
gsea_chem_df <- as.data.frame(gsea_chem) |> mutate(Condition = "CoCl2")

# Вибираємо топ-5 шляхів для кожної умови за padj
combined_res <- bind_rows(
  gsea_real_df |> arrange(p.adjust) |> head(5),
  gsea_chem_df |> arrange(p.adjust) |> head(5)
)

# Малюємо порівняльний dotplot
ggplot(combined_res, aes(x = Condition, y = ID, size = setSize, color = NES)) +
  geom_point() +
  scale_color_gradient2(low = "blue", high = "red", mid = "white") +
  theme_minimal() +
  labs(title = "Comparison of Enriched KEGG Pathways",
       y = "Pathway", x = "Condition")
```


While broad categories such as 'Pathways in Cancer' and various neurodegenerative disease signatures (e.g., Alzheimer's) showed high statistical significance, they were not prioritized for detailed analysis. Their enrichment is likely attributed to extensive gene overlap with core mitochondrial respiration and general survival signaling. Instead, the focus was directed toward more granular metabolic and stress-response pathways, which provide a clearer distinction between the physiological hypoxia and chemical stress phenotypes.


```{r}
# Real Hypoxia
p1_a <- gseaplot2(gsea_real, geneSetID = "KEGG_RIBOSOME", title = "Ribosome: Real Hypoxia", color = "#E41A1C")
# CoCl2
p1_b <- gseaplot2(gsea_chem, geneSetID = "KEGG_RIBOSOME", title = "Ribosome: CoCl2", color = "#377EB8")

# Об'єднуємо їх у рядок (порівняння "пліч-о-пліч")
panel_ribo <- ggarrange(as.ggplot(p1_a), as.ggplot(p1_b), ncol = 2)

# Real Hypoxia (має бути чіткий пік вгору)
p2_a <- gseaplot2(gsea_real, geneSetID = "KEGG_GLYCOLYSIS_GLUCONEOGENESIS", title = "Glycolysis: Real Hypoxia", color = "#E41A1C")
# CoCl2 (тут крива може бути набагато слабшою)
p2_b <- gseaplot2(gsea_chem, geneSetID = "KEGG_GLYCOLYSIS_GLUCONEOGENESIS", title = "Glycolysis: CoCl2", color = "#377EB8")

panel_glyc <- ggarrange(as.ggplot(p2_a), as.ggplot(p2_b), ncol = 2)

# Real Hypoxia
p3_a <- gseaplot2(gsea_real, geneSetID = "KEGG_VALINE_LEUCINE_AND_ISOLEUCINE_DEGRADATION", title = "Amino Acids: Real Hypoxia", color = "#E41A1C")
# CoCl2 (тут має бути сильний сигнал деградації)
p3_b <- gseaplot2(gsea_chem, geneSetID = "KEGG_VALINE_LEUCINE_AND_ISOLEUCINE_DEGRADATION", title = "Amino Acids: CoCl2", color = "#377EB8")

panel_amino <- ggarrange(as.ggplot(p3_a), as.ggplot(p3_b), ncol = 2)

# Об'єднуємо всі три пари в один великий графік
final_report_plot <- ggarrange(panel_ribo, panel_glyc, panel_amino, 
                               nrow = 3, labels = c("A", "B", "C"))

print(final_report_plot)
```
**Key Findings** 

* Panel A: Both conditions exhibited a robust and nearly identical enrichment of the KEGG_RIBOSOME pathway. This indicates a shared cellular strategy to manage stress by modulating protein synthesis and translational machinery.
* Panel B: A significant difference was observed in the KEGG_GLYCOLYSIS_GLUCONEOGENESIS pathway. While real hypoxia triggered a strong and coordinated activation of glycolytic genes , CoCl2 showed a markedly weaker and less organized response. This suggests that chemical induction fails to fully replicate the metabolic shift characteristic of true oxygen deprivation.
* Panel C: CoCl2 treatment specifically induced alterations in the KEGG_VALINE_LEUCINE_AND_ISOLEUCINE_DEGRADATION pathway, which remained unaffected in real hypoxia. This enrichment likely reflects a secondary "chemical footprint" related to metal-induced toxicity rather than hypoxic signaling.



