---
title: "Caki-1 RNA Seq"
self_contained: true
output:
  html_document:
    theme: lumen        
    highlight: tango     
    toc: true            
    toc_float: true
    code_folding: hide 
    number_sections: false     
    df_print: paged   
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  fig.align = "center",
  out.width = "100%"
)
```

### Introduction

```{r upload and inspect data, results='hide'}
# Library imports
library(tidyverse)
library(DESeq2)
library(EnsDb.Hsapiens.v86)
library(EnhancedVolcano)
library(dplyr)
library(ggplot2)
library(ggplotify)
library(pheatmap)
library(apeglm)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot)
library(DT)
library(patchwork)
```

```{r upload and inspect data, results='hide'}
# Read and inspect the dataset
load("rse_gene.Rdata")
rse <- rse_gene
colnames(colData(rse_gene))
head(colData(rse)[, c("run", "title")])
```

```{r}
mapping <- data.frame(
  run = c("SRR2970559", 
          "SRR2970560", 
          "SRR2970562", 
          "SRR2970566", 
          "SRR2970569", 
          "SRR2970570", 
          "SRR2970573", 
          "SRR2970575", 
          "SRR2970578"),
  cond = c(rep("Untreated", 3), rep("CoCl2", 3), rep("Hypoxia", 3))
)
rse$condition <- factor(rep(c("Untreated", "CoCl2", "Hypoxia"), each = 3), 
                        levels = c("Untreated", "Hypoxia", "CoCl2"))
as.data.frame(colData(rse)[, c("run", "condition")])
```

```{r upload and inspect data, results='hide'}
assay(rse)[1:5, 1:5]    # check the genes designation
# make the gene names readable for annotation
rownames(rse) <- gsub(pattern = "\\..+", replacement = "", x = rownames(rse))
assay(rse)[1:5, 1:5]
all(assay(rse) == floor(assay(rse)))    # check if counts are integers
all(rownames(colData(rse)) == colnames(assay(rse)))    # check if the metadata match count matrix
colData(rse) <- colData(rse)[, c("sample", "run", "condition")] # select data for analysis
rse$condition <- relevel(rse$condition, ref = "Untreated") # assign the control factor explicitly
```

### QC and Normalization

#### PCA Plot

```{r normalization&PCA}
# Normalization
dds <- DESeqDataSet(rse, design = ~condition)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
vsd <- vst(dds, blind = FALSE)

# PCA
pcaData <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(x = PC1, y = PC2, color = condition, label = name)) +
  geom_point(size = 5) +
  xlab(paste0("PC1 - ", percentVar[1], "%")) +
  ylab(paste0("PC2 - ", percentVar[2], "%")) +
  theme_classic(base_size = 18) +
  ggtitle("Caki-1") +
  theme(plot.title = element_text(size = 22, face = "bold"))
```


#### MA Plot

```{r DESeq and MA plot}
dds <- DESeq(dds)
resultsNames(dds)
res_real_hypoxia <- results(dds, contrast = c("condition", "Hypoxia", "Untreated"))
resNorm_real_hypoxia <- lfcShrink(dds = dds, 
                     coef = "condition_Hypoxia_vs_Untreated", 
                     type = "apeglm")

res_chemical_hypoxia <- results(dds, contrast = c("condition", "CoCl2", "Untreated"))
resNorm_chemical_hypoxia <- lfcShrink(dds = dds, 
                     coef = "condition_CoCl2_vs_Untreated", 
                     type = "apeglm")

par(mfrow = c(1, 2))
common_ylim <- c(-2.5, 2.5)
plotMA(resNorm_real_hypoxia, main = "Real Hypoxia", ylim = common_ylim)
plotMA(resNorm_chemical_hypoxia, main = "Chemical Hypoxia (CoCl2)", ylim = common_ylim)
par(mfrow = c(1, 1))
```
The two plots are nearly identical with just a few exceptions: in CoCl2 treated samples the number of upregulated genes is even higher and they have higher LFC value, however it does not indicate that this treatment mimics the real hypoxia conditions.

### Differentially expressed genes

#### Volcano plot

```{r volcano plot}
# Global settings
p_adj_cut <- 0.01
lfc_cut <- 1
base_mean_cut <- 20

# Annotation
ens2sym <- AnnotationDbi::select(EnsDb.Hsapiens.v86,
  keys = keys(EnsDb.Hsapiens.v86),
  columns = c("SYMBOL")
)

resdf_real_hypoxia <- as.data.frame(resNorm_real_hypoxia) |>
  rownames_to_column() |>
  mutate(GENEID = gsub(rowname, pattern = "\\..+", replacement = "")) |>
  dplyr::select(-rowname) |>
  inner_join(y = ens2sym, by = "GENEID")

resdf_chemical_hypoxia <- as.data.frame(resNorm_chemical_hypoxia) |>
  rownames_to_column() |>
  mutate(GENEID = gsub(rowname, pattern = "\\..+", replacement = "")) |>
  dplyr::select(-rowname) |>
  inner_join(y = ens2sym, by = "GENEID")
```


```{r volcano plot}


# Use legend and subtitle from only one plot
# Real Hypoxia
p1 <- EnhancedVolcano(resdf_real_hypoxia,
  lab = resdf_real_hypoxia$SYMBOL,
  x = "log2FoldChange", y = "padj",
  pCutoff = p_adj_cut, FCcutoff = 2,
  title = "Real Hypoxia",
  subtitle = NULL, 
  caption = NULL,  
  gridlines.major = FALSE, gridlines.minor = FALSE,
  border = "full",
  legendPosition = "none" 
)

# Chemical Hypoxia
p2 <- EnhancedVolcano(resdf_chemical_hypoxia,
  lab = resdf_chemical_hypoxia$SYMBOL,
  x = "log2FoldChange", y = "padj",
  pCutoff = p_adj_cut, FCcutoff = 2,
  title = "Chemical Hypoxia (CoCl2)",
  subtitle = NULL,
  gridlines.major = FALSE, gridlines.minor = FALSE,
  border = "full"
)

# Join with one legend
(p1 | p2) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```
It is now even more obvious that chemical and real hypoxia are not the same as the DEG names are different and the volcano profiles are also different, but still both effect gene expression levels.
#### Top 20 Significant DEGs

```{r}
# find top DEGs for both conditions
top_real_names <- sig_res_real |> arrange(padj) |> head(20) |> pull(SYMBOL)
top_chem_names <- sig_res_chemical |> arrange(padj) |> head(20) |> pull(SYMBOL)

# find common unique top DEGs
all_top_genes <- unique(c(top_real_names, top_chem_names))

# join the top genes into common table to compare
comparison_table <- resdf_real_hypoxia |>
  filter(SYMBOL %in% all_top_genes) |>
  dplyr::select(SYMBOL, LFC_Real = log2FoldChange, padj_Real = padj) |>
  inner_join(
    resdf_chemical_hypoxia |> dplyr::select(SYMBOL, LFC_Chem = log2FoldChange, padj_Chem = padj),
    by = "SYMBOL"
  )

# display one table
datatable(comparison_table)
```

Table contains 37 genes, which means that these conditions only share 3 genes as top DEGs.
The LFC for DEGs also differ.











