---
title: "Gapminder analysis"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # show code
  warning = FALSE, # hide warnings
  message = FALSE, # hide messages
  fig.align = "center" # alignment of the figures
)
```
## Technical Framework

### Core libraries

```{r packages}
library(tidyverse)  # Core suite for data manipulation and visualization
library(plotly)     # Dynamic and interactive web-based plots
library(broom)      # Converts statistical models into tidy data frames
library(knitr)      # General-purpose tool for tables
library(kableExtra) # Advanced styling tables

```
### Dataset acquisition

To simply read the file R has in-built tools, but it can result in a simple data.frame.
`read_csv` returns a `tibble` (a compact style with auto-defined data types and is faster with the big files)

```{r data read}
original_data <- read_csv("data/gapminder_clean.csv")
```
## Global wealth vs. carbon footprint

### Snapshot of 1962: The economic-environmental link

Data for 1962 was extracted and log-transformed to manage crowded areas on the plot. 
To ensure convenience, axis ticks display the raw data values.

```{r plot 1(Q1)}
# Data preparation
filtered_data <- original_data |>
  filter(Year == 1962) # Extract the subset for 1962

# Plotting
plot_1 <- ggplot(
  filtered_data,
  aes(
    x = gdpPercap,
    y = `CO2 emissions (metric tons per capita)`, # Configure the plot axes
    text = paste(
      "Country:", `Country Name`, # Define hover tooltip content
      "<br>GDP per capita:", gdpPercap,
      "<br>CO2:", `CO2 emissions (metric tons per capita)`
    )
  )
) +
  geom_point(alpha = 0.6) + # Add scatterplot with semitransparent dots
  scale_x_log10() + # Functional requirement: log scale for GDP
  scale_y_log10() + # Functional requirement: log scale for CO2
  theme_classic()

# Plot appearance
ggplotly(plot_1, tooltip = "text") # Make interactive
```
A specific data point represents a suspicious outlier, warranting further inspection of the underlying data.

A review of the literature reveals the following findings.
The spike in Kuwait’s CO₂ emissions in 1962—which reached world-record levels on a per capita basis—was primarily driven by the massive flaring of associated petroleum gas.
During this period, the country rapidly expanded oil production following its independence (1961), but lacked the infrastructure to capture or utilize the natural gas released during extraction, resulting in it being burned off as waste.
Because these vast industrial emissions were divided by a relatively small population (358,266 people at the time - according to this data set), the mathematical result was an extreme statistical anomaly that far exceeded the per capita output of even the most industrialized Western nations.

### Quantifying the correlation

To quantify the strength and direction of the linear relationship between GDP per capita and CO₂ emissions, a Pearson correlation test was conducted.

```{r correlation-analysis}
# Pearson correlation test
cor_results <- cor.test(
  filtered_data$gdpPercap,
  filtered_data$`CO2 emissions (metric tons per capita)`
)

# Transforming to table
correlation_table <- tidy(cor_results) |>
  mutate(p.value = formatC(p.value, format = "e", digits = 2)) |>
  kable(digits = 3, caption = "Pearson correlation test results")

# Show results
correlation_table
```
The high correlation between GDP per capita and CO₂ emissions - `r round(cor_results$estimate, 3)`
with a low p-value of `r format.pval(cor_results$p.value, digits = 3, eps = .Machine$double.eps)` indicate a statistically significant and strong positive correlation between the variables.

The objective is to identify the specific year in which the correlation between GDP per capita and CO₂ emissions reached its peak. While `cor.test()` provides a more comprehensive statistical output, the simpler `cor()` function is preferred for grouped data processing to efficiently compute and compare coefficients across multiple years.

```{r correlation in groups }
# Correlation test in groups
cor_by_year <- original_data |>
  group_by(Year) |>
  summarise(
    correlation = cor(
      gdpPercap,
      `CO2 emissions (metric tons per capita)`,
      use = "complete.obs"
    ),
    .groups = "drop"
  )
# Find out which year has the strongest correlation
max_cor_year <- cor_by_year |>
  filter(correlation == max(correlation)) |>
  pull(Year)
```

So, the year with the strongest correlation is `r max_cor_year`.
Now let's focus and discover this period of time by making simple visualizations

```{r visualization of the cor.test result year}
# Data Preparation: filtering out entities without continent information
data_max_cor <- original_data |>
  filter(Year == max_cor_year) |>
  filter(!is.na(continent))

# Plot setup
plot_2 <- ggplot(
  data_max_cor,
  aes(
    x = gdpPercap,
    y = `CO2 emissions (metric tons per capita)`,
    color = continent, # Core: add a third dimension (continents)
    size = pop, # Core: add a fourth dimension (population)
    text = paste(
      "Country:", `Country Name`, # Define hover tooltip content
      "<br>GDP per capita:", gdpPercap,
      "<br>CO2:", round(`CO2 emissions (metric tons per capita)`, 2), "t/capita",
      "<br>Population:", pop
    )
  )
) +
  # Add scatterplot with semitransparent dots (na.rm to avoid warnings)
  geom_point(alpha = 0.7, na.rm = TRUE) +
  scale_x_log10() +
  scale_y_log10() +
  scale_size_continuous(guide = "none") + # Hide size legend for better readability
  theme_classic() +
  labs( # Guides
    title = paste("Global correlation between GDP and CO2 emissions in", max_cor_year),
    color = "Continent"
  )
# Make interactive
ggplotly(plot_2, tooltip = "text")
```

The visualization above shows the relationship for `r max_cor_year`. Note that the size of each circle reflects the country's population, though the size legend has been omitted for clarity.

## Exploratory data analysis (EDA)

### Continental disparities in energy consumption

Question 1. What is the relationship between continent and Energy use (kg of oil equivalent per capita)?

As the independent variable (continent) is categorical, the analysis focuses on identifying significant differences between groups. The dependent variable is quantitative (energy use), and the number of groups exceeds two. Given that we are comparing a single dependent variable across these categories, a one-way ANOVA is the most appropriate statistical test. 
```{r ANOVA}
# ANOVA test
res_anova <- aov(`Energy use (kg of oil equivalent per capita)` ~ continent, data = original_data)
# Transforming to table
res_anova_t <- tidy(res_anova) |>
  mutate(
    statistic = ifelse(is.na(statistic), "", round(statistic, 2)), # Hide NA
    p.value = ifelse(is.na(p.value), "", formatC(p.value, format = "e", digits = 2))
  ) |>
  kable(digits = 3, caption = "ANOVA results")
# Show results
res_anova_t
```
To identify which specific continents differ from one another, a Tukey’s honest significance difference (HSD) test was performed as a post-hoc analysis following the ANOVA.
```{r Tukey}
# Tukey's honest significance test (table transformed)
tukey_data <- TukeyHSD(res_anova) |>
  tidy()
tukey_table <- tukey_data |>
  mutate(adj.p.value = formatC(adj.p.value, format = "e", digits = 2)) |>
  kable(digits = 3, caption = "Tukey results")
# Show results
tukey_table

# Visualize the Tukey test results
plot(TukeyHSD(res_anova))
```
The ANOVA test indicates that the continent is a critical factor for energy consumption levels `r format.pval(res_anova$p.value, digits = 3, eps = .Machine$double.eps)`. A follow-up Tukey test clarified that almost all continents differ significantly from one another, with the exceptions of the Oceania–Europe and Asia–Americas pairs, where energy consumption is statistically similar. However, the ANOVA calculation revealed a substantial amount of excluded data due to missing values. Visual analysis of the missing data structure shows a higher concentration of gaps within the African continent. This likely influences the statistical test results and must be taken into account when interpreting the findings.
```{r missing data plots}
original_data |>
  # Filter out NAs
  filter(!is.na(continent)) |>
  group_by(continent) |>
  summarise(
    pct_missing = mean(is.na(`Energy use (kg of oil equivalent per capita)`)) * 100
  ) |>
  # Building a plot
  ggplot(aes(x = reorder(continent, -pct_missing), y = pct_missing, fill = continent)) +
  geom_col() +
  # visuals
  scale_fill_viridis_d(option = "G", begin = 0.2, end = 0.9) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(
    title = "Percentage of missing energy consumption Data",
    subtitle = "Analysis by continent for the selected period",
    x = "Continent",
    y = "% of missing values"
  )

```

### Comparative analysis of import reliance: Europe vs. Asia

Question 2.Is there a significant difference between Europe and Asia with respect to 'Imports of goods and services (% of GDP)' in the years after 1990?

This is a classic statistical problem: comparing two groups to determine whether a significant difference exists between them. Our grouping (independent) variable is categorical with exactly two levels. The dependent variable is continuous (interval/ratio).

Following the standard statistical flowchart, the Independent Samples T-test was selected. A key prerequisite for this test is that the data within both groups must follow a normal distribution. To verify this assumption, the Shapiro-Wilk test can be employed. If the p-value is less than 0.05, the null hypothesis of normality is rejected, indicating that the distribution cannot be considered normal.

```{r Shapiro-Wilk for Europe and Asia}
# Preparing the data
comparison_df <- original_data |>
  filter(Year > 1990) |>
  filter(continent %in% c("Europe", "Asia"))

# # Shapiro-Wilk test results for visualization labels
p_labels <- comparison_df |>
  group_by(continent) |>
  summarise(
    p_val = shapiro.test(`Imports of goods and services (% of GDP)`)$p.value,
    # Adding a p-value to each facet
    label = paste0("p = ", format(p_val, digits = 3))
  )

# Generate Q-Q plot to check normality
ggplot(comparison_df, aes(sample = `Imports of goods and services (% of GDP)`)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  # Annotations
  geom_text(
    data = p_labels, aes(x = -1.5, y = 150, label = label),
    inherit.aes = FALSE, color = "blue", size = 4, hjust = 0
  ) +
  facet_wrap(~continent) +
  labs(
    title = "Shapiro-Wilk normality test: Imports of goods and services (% of GDP) ",
    x = "Theoretical quantiles",
    y = "Sample quantiles (%GDP)"
  ) +
  theme_bw()
```
Both data-sets do not follow a normal distribution (deviations from the reference line are clearly visible at both the lower and upper quantiles), and the p-values are significantly lower than 0.05. Consequently, the non-parametric Mann-Whitney U test (Wilcoxon rank-sum test for independent samples) should be applied to compare the two continents.

```{r Mann-Whitney U test}
final_test <- wilcox.test(
  `Imports of goods and services (% of GDP)` ~
    continent,
  data = comparison_df
)
final_test_table <- final_test |> # Transform to the table
  tidy() |>
  kable(digits = 3, caption = "Mann-Whitney U test results")
# Show results
final_test_table
```

The null hypothesis ($H_0$) for the Mann-Whitney U test states that there is no difference between the distributions of imports in Asia and Europe. Given the p-value of 0.787 (considerably higher than the 0.05 threshold), we fail to reject $H_0$, indicating that any observed differences between the two continents are likely due to chance.

```{r boxplot fo import in Europe and Asia}
ggplot(comparison_df, aes(x = continent, y = `Imports of goods and services (% of GDP)`, fill = continent)) +
  geom_boxplot(alpha = 0.7) +
  # Add jitter for clarity
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(
    title = "Import comparison: Europe vs Asia",
    subtitle = paste("Wilcoxon test p-value =", round(final_test$p.value, 4)),
    x = "Continent",
    y = "Import (% GDP)",
    caption = "Data after 1990"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Accent")
```
This boxplot confirms the initial hypothesis: Asia has a large number of extreme outliers compared to Europe, which could cause the arithmetic means to differ. However, a comparison of the medians demonstrates just how similar these two distributions actually are.


