---
title: "Gapminder analysis"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # show code
  warning = FALSE, # hide warnings
  message = FALSE, # hide messages
  fig.align = "center" # alignment of the figures
)
```
## Technical Framework

### Core libraries

```{r packages}
library(tidyverse)   # Core suite for data manipulation and visualization
library(plotly)      # Dynamic and interactive web-based plots
library(broom)       # Converts statistical models into tidy data frames
library(knitr)       # General-purpose tool for tables
library(kableExtra)  # Advanced styling tables


```
### Dataset acquisition

To simply read the file R has in-built tools, but it can result in a simple data.frame.
`read_csv` returns a `tibble` (a compact style with auto-defined data types and is faster with the big files)

```{r data read}
original_data <- read_csv("data/gapminder_clean.csv")
```
## Global wealth vs. carbon footprint

### Snapshot of 1962: The economic-environmental link

Data for 1962 was extracted and log-transformed to manage crowded areas on the plot. 
To ensure convenience, axis ticks display the raw data values.

```{r plot 1(Q1)}
# Data preparation
filtered_data <- original_data |>
  filter(Year == 1962) # Extract the subset for 1962

# Plotting
plot_1 <- ggplot(
  filtered_data,
  aes(
    x = gdpPercap,
    y = `CO2 emissions (metric tons per capita)`, # Configure the plot axes
    text = paste(
      "Country:", `Country Name`, # Define hover tooltip content
      "<br>GDP per capita:", gdpPercap,
      "<br>CO2:", `CO2 emissions (metric tons per capita)`
    )
  )
) +
  geom_point(alpha = 0.6) + # Add scatterplot with semitransparent dots
  scale_x_log10() + # Functional requirement: log scale for GDP
  scale_y_log10() + # Functional requirement: log scale for CO2
  theme_classic()

# Plot appearance
ggplotly(plot_1, tooltip = "text") # Make interactive
```
A specific data point represents a suspicious outlier, warranting further inspection of the underlying data.

A review of the literature reveals the following findings.
The spike in Kuwait’s CO₂ emissions in 1962—which reached world-record levels on a per capita basis—was primarily driven by the massive flaring of associated petroleum gas.
During this period, the country rapidly expanded oil production following its independence (1961), but lacked the infrastructure to capture or utilize the natural gas released during extraction, resulting in it being burned off as waste.
Because these vast industrial emissions were divided by a relatively small population (358,266 people at the time - according to this data set), the mathematical result was an extreme statistical anomaly that far exceeded the per capita output of even the most industrialized Western nations.

### Quantifying the correlation

To quantify the strength and direction of the linear relationship between GDP per capita and CO₂ emissions, a Pearson correlation test was conducted.

```{r correlation-analysis}
# Pearson correlation test
cor_results <- cor.test(
  filtered_data$gdpPercap,
  filtered_data$`CO2 emissions (metric tons per capita)`
)

# Transforming to table
correlation_table <- tidy(cor_results) |>
  mutate(p.value = formatC(p.value, format = "e", digits = 2)) |>
  kable(digits = 3, caption = "Pearson correlation test results")

# Show results
correlation_table
```

