---
title: "Single-Cell RNA-Seq"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(Seurat)
library(tidyverse)
library(ggpubr)
library(Matrix)
library(ggplot2)
library(patchwork)
library(SingleR)
library(celldex)
library(SingleCellExperiment)
```

```{r}
if (!file.exists("melanoma_matrix.tar.gz")) {
  download.file(
    url = "https://cf.10xgenomics.com/samples/cell-exp/8.0.0/10k_Human_DTC_Melanoma_3p_nextgem_10k_Human_DTC_Melanoma_3p_nextgem/10k_Human_DTC_Melanoma_3p_nextgem_10k_Human_DTC_Melanoma_3p_nextgem_count_sample_filtered_feature_bc_matrix.tar.gz",
    destfile = "melanoma_matrix.tar.gz",
    method = "curl"
  )
}
```


```{r}
if (!dir.exists("sample_filtered_feature_bc_matrix")) {
  untar("melanoma_matrix.tar.gz")
}
list.dirs(recursive = FALSE)
```

```{r}
mat <- Read10X(data.dir = "sample_filtered_feature_bc_matrix")
melanoma_raw <- CreateSeuratObject(counts = mat, project = "Melanoma_10k")
```

```{r}
View(mat)
View(melanoma_raw)
```

```{r}
melanoma_raw <- melanoma_raw |> 
  PercentageFeatureSet(pattern = "^MT-", col.name = "percent.mt")

VlnPlot(melanoma_raw, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
FeatureScatter(melanoma_raw, feature1 = "nCount_RNA", feature2 = "percent.mt")
```

```{r}
melanoma <- melanoma_raw |> 
  subset(subset = nFeature_RNA > 500 & nFeature_RNA < 6000 & percent.mt < 10) |>  
  NormalizeData() |> 
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000) |> 
  ScaleData()

# check the results
melanoma |> 
  FeatureScatter(feature1 = "nCount_RNA", feature2 = "percent.mt") +
  theme_bw()
```
#### PCA
```{r}
melanoma <- RunPCA(melanoma)

DimPlot(melanoma)

# What are the genes that determine PC1 and PC2?
VizDimLoadings(melanoma, dims = 1:2)
```


```{r}
FeaturePlot(melanoma, features = c("S100A1", "CST3", "EDNRB"))

FeaturePlot(melanoma, features = c("LYZ", "TYROBP"))

# Select number of PCs
ElbowPlot(melanoma, ndims = 50)
```
NOTES: 
1.Endothelin receptor B (ETRB or EDNRB) is overexpressed in most human melanomas and is proposed to provide a marker of melanoma progression.
2. Overexpression of S100A1 is associated with melanoma, breast cancer, ovarian cancer, and endometrial carcinoma (similar to S100A1 expression profile).
3. Factors upregulated CysC secretion by melanoma. Similarly, melanoma-derived factors upregulated CysC secretion by microglia.
4. TYROBP is a prognostic marker in Skin cutaneous melanoma/ Cell type enhanced (Hofbauer cells, Kupffer cells, Neutrophils, monocytes, Macrophages, cDC, pDCs, NK-cells, Mast cells)

```{r}
melanoma <- melanoma |> 
  FindNeighbors(dims = 1:25) |> 
  FindClusters(resolution = 0.5) |>  
  RunUMAP(dims = 1:25)

DimPlot(melanoma, reduction = "umap", label = TRUE) + NoLegend()
```


## Marker annotation
```{r}
melanoma.markers <- FindAllMarkers(melanoma,
                               only.pos = TRUE, 
                               min.pct = 0.25, 
                               logfc.threshold = 0.5)
```
Step 1. Prior knowledge markers

```{r}
top5_markers <- melanoma.markers |> 
  group_by(cluster) |> 
  slice_max(n = 5, order_by = avg_log2FC)

View(top5_markers)
# Look for "unique only/mostly" markers 
for(i in unique(top5_markers$cluster)) {
  genes_to_plot <- top5_markers %>% filter(cluster == i) |>  pull(gene)
  
  p <- FeaturePlot(melanoma, features = genes_to_plot, ncol = 3) + 
    plot_annotation(title = paste("Top markers for Cluster", i))
  
  print(p)
}

```


```{r}
marker_list_melanoma <- list(
  'Neurons/tumor' = c("SYN3", "NETO1"),
  'T cells' = c("CD40LG"),
  'CTLA' = c("BTNL9", "CD80"),
  'Granulosa cells/pro-inflammatory' = c("ANKS1B", "ICA1", "TOX2"),
  'NK' = c("NKG7"),
  'Macrophages' = c("CD163"),
  'Plasma cells' = c("IGHG3", "IGHGP"),
  'Endothelial/melanocytes' = c("EDN3"))
marker_vln <- function(srt, marker_list, group_name) {
  genes <- marker_list[[group_name]]
  vln <- VlnPlot(srt, features = genes, ncol = length(genes), pt.size = 0) +
    patchwork::plot_annotation(title = group_name)
  return(vln)
}

vlnList <- lapply(names(marker_list_melanoma), marker_vln, 
                  srt = melanoma, marker_list = marker_list_melanoma)
final_ga <- ggarrange(plotlist = vlnList, ncol = 1)

ggsave(final_ga, filename = "melanoma_validation_violins.png", height = 25, width = 15)
```

Step 2. Semi-automated annotation

```{r, fig.height=10}
bped.se <- BlueprintEncodeData()
sceP <- as.SingleCellExperiment(melanoma)
pred.bped.main <- SingleR(test = sceP, 
                          ref = bped.se, 
                          labels = bped.se$label.main)
plotScoreHeatmap(pred.bped.main,
                 max.labels = 8,
                 clusters = melanoma$seurat_clusters,
                 order.by = "clusters",
                 show_colnames = FALSE)
```
From this perspective it is quite clear that claster 11 contains melanocytes (tumor) and all other clasters are the immune microenvironment. To get a deeper understanding of what is going on in the immune microenvironment the MonacoImmuneData can be utilized.

```{r, fig.height=15 }
mimd.se <- MonacoImmuneData()

sceP <- as.SingleCellExperiment(melanoma)
pred.mimd.main <- SingleR(test = sceP, 
                          ref = mimd.se, 
                          labels = mimd.se$label.main)
pred.mimd.fine <- SingleR(test = sceP, 
                          ref = mimd.se, 
                          labels = mimd.se$label.fine)
plotScoreHeatmap(pred.mimd.main,
                 max.labels = 20,
                 clusters = melanoma$seurat_clusters,
                 order.by = "clusters",
                 show_colnames = FALSE)
plotScoreHeatmap(pred.mimd.fine,
                 max.labels = 20,
                 clusters = melanoma$seurat_clusters,
                 order.by = "clusters",
                 show_colnames = FALSE)

```
Some of the "fine" labels seem to be robust, but some are not.


```{r}
### 1. Re-train SingleR (Optimizing for available genes)
# This step ensures SingleR only uses genes present in your specific dataset
my_genes <- rownames(sceP)

trained_mimd_main <- trainSingleR(mimd.se, 
                                  labels = mimd.se$label.main, 
                                  test.genes = my_genes)

trained_mimd_fine <- trainSingleR(mimd.se, 
                                  labels = mimd.se$label.fine, 
                                  test.genes = my_genes)

# Combine Main and Fine level predictions for higher accuracy
pred.comb <- combineRecomputedResults(
  results = list(Main = pred.mimd.main, Fine = pred.mimd.fine),
  test = sceP, 
  trained = list(Main = trained_mimd_main, Fine = trained_mimd_fine),
  assay.type.test = "logcounts"
)

### 2. Diagnostics & Confidence Assessment
# Check which markers were used and assess assignment confidence
metadata(pred.mimd.fine)$common.markers
pred.mimd.fine$clusters <- melanoma$seurat_clusters

# Visualizing score distribution (how clear the "winner" is for each cell)
plotScoreDistribution(pred.mimd.fine)

### 3. Final Integration & Validation
# Map the high-confidence (pruned) labels back to the Seurat object
melanoma$predicted_id <- pred.comb$pruned.labels

# Final summary heatmap of combined results
plotScoreHeatmap(pred.comb,
                 max.labels = 20,
                 clusters = melanoma$seurat_clusters,
                 order.by = "clusters",
                 show_colnames = FALSE,
                 main = "Final Combined Annotation")

# Cross-tabulation to compare Clusters vs. Predicted IDs
table(melanoma$predicted_id, melanoma$seurat_clusters)

```
```{r}
Idents(melanoma) <- melanoma$seurat_clusters
new_ids <- c(
  "0" = "Naive B cells",
  "1" = "Naive CD4 T cells*",
  "2" = "Non-switched memory B cells",
  "3" = "Naive CD4 T cells*",
  "4" = "Switched memory B cells",
  "5" = "T regulatory cells",
  "6" = "Naive B cells",
  "7" = "Naive CD8 T cells",
  "8" = "Vd2 gd T cells",
  "9" = "Myeloid dendritic",
  "10" = "B cells*",
  "11" = "Melanocytes",
  "12" = "Plasmacytoid dendritic"
)
melanoma <- RenameIdents(melanoma, new_ids)
melanoma$cell_type <- Idents(melanoma)
DimPlot(melanoma, label.size = 3, label = TRUE) + NoLegend() 
```
The T-cell, B-cell, and dendritic cell populations are clearly distinct, with the melanocyte cluster also forming a separate group. One specific B-cell cluster is particularly noteworthy as it is positioned away from the others; it expresses a set of exhausted B-cell markers absent in other clusters, potentially indicating a hidden communication axis with the melanoma. Fully identifying sub-populations remains challenging because this map captures a 'dynamic cell transformation' process. As markers gradually appear and disappear, a 'pseudotime' or trajectory analysis is required to resolve these transitions.
