---
title: "Single-Cell RNA-Seq"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r setup, include=FALSE}
library(reticulate)
library(Matrix)
library(phateR)
library(Rmagic)
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(patchwork)
library(SingleR)
library(celldex)
library(SingleCellExperiment)
library(enrichR)
library(clusterProfiler)
library(Seurat)
library(monocle3)
library(SeuratWrappers)
library(CellChat)
library(ggalluvial)
reticulate::py_require(c("magic-impute", "phate"))
reticulate::py_config()
if (!reticulate::py_module_available("phate") |
    !reticulate::py_module_available("magic")) {
  reticulate::py_install(c("magic-impute", "phate"), pip = TRUE)
}
```

```{r}
if (!file.exists("melanoma_matrix.tar.gz")) {
  download.file(
    url = "https://cf.10xgenomics.com/samples/cell-exp/8.0.0/10k_Human_DTC_Melanoma_3p_nextgem_10k_Human_DTC_Melanoma_3p_nextgem/10k_Human_DTC_Melanoma_3p_nextgem_10k_Human_DTC_Melanoma_3p_nextgem_count_sample_filtered_feature_bc_matrix.tar.gz",
    destfile = "melanoma_matrix.tar.gz",
    method = "curl"
  )
}
```


```{r}
if (!dir.exists("sample_filtered_feature_bc_matrix")) {
  untar("melanoma_matrix.tar.gz")
}
list.dirs(recursive = FALSE)
```

```{r}
mat <- Read10X(data.dir = "sample_filtered_feature_bc_matrix")
melanoma_raw <- CreateSeuratObject(counts = mat, project = "Melanoma_10k")
```

```{r}
View(mat)
View(melanoma_raw)
```

```{r}
melanoma_raw <- melanoma_raw |> 
  PercentageFeatureSet(pattern = "^MT-", col.name = "percent.mt")

VlnPlot(melanoma_raw, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
FeatureScatter(melanoma_raw, feature1 = "nCount_RNA", feature2 = "percent.mt")
```

```{r}
melanoma <- melanoma_raw |> 
  subset(subset = nFeature_RNA > 500 & nFeature_RNA < 6000 & percent.mt < 10) |>  
  NormalizeData() |> 
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000) |> 
  ScaleData()

# check the results
melanoma |> 
  FeatureScatter(feature1 = "nCount_RNA", feature2 = "percent.mt") +
  theme_bw()
```
#### PCA
```{r}
melanoma <- RunPCA(melanoma)

DimPlot(melanoma)

# What are the genes that determine PC1 and PC2?
VizDimLoadings(melanoma, dims = 1:2)
```


```{r}
FeaturePlot(melanoma, features = c("S100A1", "CST3", "EDNRB"))

FeaturePlot(melanoma, features = c("LYZ", "TYROBP"))

# Select number of PCs
ElbowPlot(melanoma, ndims = 50)
```
NOTES: 
1.Endothelin receptor B (ETRB or EDNRB) is overexpressed in most human melanomas and is proposed to provide a marker of melanoma progression.
2. Overexpression of S100A1 is associated with melanoma, breast cancer, ovarian cancer, and endometrial carcinoma (similar to S100A1 expression profile).
3. Factors upregulated CysC secretion by melanoma. Similarly, melanoma-derived factors upregulated CysC secretion by microglia.
4. TYROBP is a prognostic marker in Skin cutaneous melanoma/ Cell type enhanced (Hofbauer cells, Kupffer cells, Neutrophils, monocytes, Macrophages, cDC, pDCs, NK-cells, Mast cells)

```{r}
melanoma <- melanoma |> 
  FindNeighbors(dims = 1:25) |> 
  FindClusters(resolution = 0.5) |>  
  RunUMAP(dims = 1:25)

DimPlot(melanoma, reduction = "umap", label = TRUE) + NoLegend()
```


## Marker annotation
```{r}
melanoma.markers <- FindAllMarkers(melanoma,
                               only.pos = TRUE, 
                               min.pct = 0.25, 
                               logfc.threshold = 0.5)
```
Step 1. Prior knowledge markers

```{r}
top5_markers <- melanoma.markers |> 
  group_by(cluster) |> 
  slice_max(n = 5, order_by = avg_log2FC)

View(top5_markers)
# Look for "unique only/mostly" markers 
for(i in unique(top5_markers$cluster)) {
  genes_to_plot <- top5_markers %>% filter(cluster == i) |>  pull(gene)
  
  p <- FeaturePlot(melanoma, features = genes_to_plot, ncol = 3) + 
    plot_annotation(title = paste("Top markers for Cluster", i))
  
  print(p)
}

```


```{r}
marker_list_melanoma <- list(
  'Neurons/tumor' = c("SYN3", "NETO1"),
  'T cells' = c("CD40LG"),
  'CTLA' = c("BTNL9", "CD80"),
  'Granulosa cells/pro-inflammatory' = c("ANKS1B", "ICA1", "TOX2"),
  'NK' = c("NKG7"),
  'Macrophages' = c("CD163"),
  'Plasma cells' = c("IGHG3", "IGHGP"),
  'Endothelial/melanocytes' = c("EDN3"))
marker_vln <- function(srt, marker_list, group_name) {
  genes <- marker_list[[group_name]]
  vln <- VlnPlot(srt, features = genes, ncol = length(genes), pt.size = 0) +
    patchwork::plot_annotation(title = group_name)
  return(vln)
}

vlnList <- lapply(names(marker_list_melanoma), marker_vln, 
                  srt = melanoma, marker_list = marker_list_melanoma)
final_ga <- ggarrange(plotlist = vlnList, ncol = 1)

ggsave(final_ga, filename = "melanoma_validation_violins.png", height = 25, width = 15)
```

Step 2. Semi-automated annotation

```{r, fig.height=10}
bped.se <- BlueprintEncodeData()
sceP <- as.SingleCellExperiment(melanoma)
pred.bped.main <- SingleR(test = sceP, 
                          ref = bped.se, 
                          labels = bped.se$label.main)
plotScoreHeatmap(pred.bped.main,
                 max.labels = 8,
                 clusters = melanoma$seurat_clusters,
                 order.by = "clusters",
                 show_colnames = FALSE)
```
From this perspective it is quite clear that cluster 11 contains melanocytes (tumor) and all other clusters are the immune microenvironment. To get a deeper understanding of what is going on in the immune microenvironment the MonacoImmuneData can be utilized.

```{r, fig.height=15 }
mimd.se <- MonacoImmuneData()

sceP <- as.SingleCellExperiment(melanoma)
pred.mimd.main <- SingleR(test = sceP, 
                          ref = mimd.se, 
                          labels = mimd.se$label.main)
pred.mimd.fine <- SingleR(test = sceP, 
                          ref = mimd.se, 
                          labels = mimd.se$label.fine)
plotScoreHeatmap(pred.mimd.main,
                 max.labels = 20,
                 clusters = melanoma$seurat_clusters,
                 order.by = "clusters",
                 show_colnames = FALSE)
plotScoreHeatmap(pred.mimd.fine,
                 max.labels = 20,
                 clusters = melanoma$seurat_clusters,
                 order.by = "clusters",
                 show_colnames = FALSE)

```
Some of the "fine" labels seem to be robust, but some are not.


```{r}
### 1. Re-train SingleR (Optimizing for available genes)
# This step ensures SingleR only uses genes present in your specific dataset
my_genes <- rownames(sceP)

trained_mimd_main <- trainSingleR(mimd.se, 
                                  labels = mimd.se$label.main, 
                                  test.genes = my_genes)

trained_mimd_fine <- trainSingleR(mimd.se, 
                                  labels = mimd.se$label.fine, 
                                  test.genes = my_genes)

# Combine Main and Fine level predictions for higher accuracy
pred.comb <- combineRecomputedResults(
  results = list(Main = pred.mimd.main, Fine = pred.mimd.fine),
  test = sceP, 
  trained = list(Main = trained_mimd_main, Fine = trained_mimd_fine),
  assay.type.test = "logcounts"
)

### 2. Diagnostics & Confidence Assessment
# Check which markers were used and assess assignment confidence
metadata(pred.mimd.fine)$common.markers
pred.mimd.fine$clusters <- melanoma$seurat_clusters

# Visualizing score distribution (how clear the "winner" is for each cell)
plotScoreDistribution(pred.mimd.fine)

### 3. Final Integration & Validation
# Map the high-confidence (pruned) labels back to the Seurat object
melanoma$predicted_id <- pred.comb$pruned.labels

# Final summary heatmap of combined results
plotScoreHeatmap(pred.comb,
                 max.labels = 20,
                 clusters = melanoma$seurat_clusters,
                 order.by = "clusters",
                 show_colnames = FALSE,
                 main = "Final Combined Annotation")

# Cross-tabulation to compare Clusters vs. Predicted IDs
table(melanoma$predicted_id, melanoma$seurat_clusters)

```

Some Identities are duplicates, it will be probably explained by metabolism analysis later.
```{r}
Idents(melanoma) <- melanoma$seurat_clusters
new_ids <- c(
  "0" = "Naive B cells",
  "1" = "Naive CD4 T cells",
  "2" = "Non-switched memory B cells",
  "3" = "Naive CD4 T cells*",
  "4" = "Switched memory B cells",
  "5" = "T regulatory cells",
  "6" = "Naive B cells*",
  "7" = "Naive CD8 T cells",
  "8" = "Vd2 gd T cells",
  "9" = "Myeloid dendritic",
  "10" = "B cells*",
  "11" = "Melanocytes",
  "12" = "Plasmacytoid dendritic"
)
melanoma <- RenameIdents(melanoma, new_ids)
melanoma$cell_type <- Idents(melanoma)
DimPlot(melanoma, label.size = 3, label = TRUE) + NoLegend() 
```
The T-cell, B-cell, and dendritic cell populations are clearly distinct, with the melanocyte cluster also forming a separate group. One specific B-cell cluster is particularly noteworthy as it is positioned away from the others; it expresses a set of exhausted B-cell markers absent in other clusters, potentially indicating a hidden communication axis with the melanoma. Fully identifying sub-populations remains challenging because this map captures a 'dynamic cell transformation' process. As markers gradually appear and disappear, a 'pseudotime' or trajectory analysis is required to resolve these transitions.

```{r}
s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

melanoma <- CellCycleScoring(melanoma, 
                             s.features = s.genes, 
                             g2m.features = g2m.genes, 
                             set.ident = TRUE)

DimPlot(melanoma, reduction = "umap", group.by = "Phase")
```
1. Clusters are defined by cell type, not cycle artifacts.
2. Tumor cells are predominantly in G1, indicating a quiescent state at sampling.
3. The distinct B-cell* cluster suggests a specific functional activation separate from naive/memory pools.

PHATE + MAGIC 

```{r}
input_data_t <- t(LayerData(melanoma, assay = "RNA", layer = "data"))

data_MAGIC <- magic(input_data_t, knn = 15, seed = 42)
phate_res <- phate(data_MAGIC, knn = 20, t = "auto")
phate_embeddings_raw <- as.matrix(phate_res$embedding)
rownames(phate_embeddings_raw) <- rownames(input_data_t)
final_mat <- matrix(NA, nrow = ncol(melanoma), ncol = 2)
rownames(final_mat) <- colnames(melanoma)
colnames(final_mat) <- c("PHATEMAGIC_1", "PHATEMAGIC_2")
common <- intersect(rownames(final_mat), rownames(phate_embeddings_raw))
final_mat[common, ] <- phate_embeddings_raw[common, ]

melanoma[["phate_magic"]] <- CreateDimReducObject(
  embeddings = final_mat, 
  key = "PHATEMAGIC_", 
  assay = "RNA"
)
my_cols <- c(
  "Naive B cells" = "#1F78B4",        
  "B cells*" = "#A6CEE3",             
  "Switched memory B cells" = "#33A02C", 
  "Non-switched memory B cells" = "#B2DF8A",
  "Naive B cells*" = "#1c9099",
  "Naive CD4 T cells" = "#E31A1C",
  "Naive CD4 T cells*" = "#800026",
  "Naive CD8 T cells" = "#FB9A99",    
  "T regulatory cells" = "#FF7F00",   
  "Vd2 gd T cells" = "#FDBF6F",      
  "Melanocytes" = "#CAB2D6",          
  "Myeloid dendritic" = "#6A3D9A",    
  "Plasmacytoid dendritic" = "#FFFF99" 
)

p1 <- DimPlot(melanoma, reduction = "umap", group.by = "cell_type", cols = my_cols, pt.size = 0.6) + labs(title = "UMAP")
p2 <- DimPlot(melanoma, reduction = "phate_magic", group.by = "cell_type", cols = my_cols, pt.size = 0.6) + labs(title = "MAGIC + PHATE")

p1 + p2

```

Pseudotime


```{r}
cds <- as.cell_data_set(melanoma)
cds <- cluster_cells(cds, reduction_method = "UMAP")
cds <- learn_graph(cds, use_partition = TRUE)
get_roots_per_partition <- function(cds) {
  parts <- unique(partitions(cds))
  root_cells <- c()
  
  for (p in parts) {
    cell_id <- which(partitions(cds) == p)[1]
    root_cells <- c(root_cells, names(cell_id))
  }
  return(root_cells)
}

cds <- order_cells(cds, 
                   reduction_method = "UMAP", 
                   root_cells = get_roots_per_partition(cds))

colData(cds)$cell_type <- melanoma$cell_type

p1 <- plot_cells(cds, 
                 color_cells_by = "cell_type", 
                 label_cell_groups = FALSE,
                 label_leaves = FALSE, 
                 graph_label_size = 0) + 
  theme_minimal()

p2 <- plot_cells(cds, 
                 color_cells_by = "pseudotime", 
                 label_cell_groups = FALSE,
                 label_leaves = FALSE, 
                 graph_label_size = 0) + 
  theme_minimal()

p1 + p2
```

Metabolism 

```{r}
sig_markers <- melanoma.markers |>  filter(p_val_adj < 0.05)

to_check <- c("Human_Gene_Atlas", "ENCODE_and_ChEA_Consensus_TFs_from_ChIP-X", 
              "KEGG_2019_Human", "MSigDB_Hallmark_2020")

pdf("All_Clusters_Enrichr_Report.pdf", width = 16, height = 10)

for (target_cluster in sort(unique(sig_markers$cluster))) {
  
  message(paste("Processing cluster:", target_cluster))

  cl_genes <- sig_markers |> 
    filter(cluster == target_cluster) |> 
    pull(gene)

  if(length(cl_genes) == 0) next

  cl_eres <- enrichr(cl_genes, databases = to_check)
  
  plotList <- lapply(names(cl_eres), function(db) {
    df <- cl_eres[[db]]
    if (is.null(df) || nrow(df) == 0) return(NULL)
    
    df |> 
      top_n(10, wt = -log10(Adjusted.P.value)) |>  
      mutate(Term = stringr::str_trunc(Term, 45)) |>   
      mutate(Term = factor(Term, levels = rev(unique(Term)))) |> 
      ggplot(aes(x = Term, y = -log10(Adjusted.P.value), fill = Combined.Score)) +
      geom_col() +
      coord_flip() +
      theme_minimal(base_size = 12) +
      labs(title = db, x = "") +
      scale_fill_gradient(low = "skyblue", high = "darkblue")
  })

  plotList <- Filter(Negate(is.null), plotList)
  
  if(length(plotList) > 0) {
    combined_plot <- cowplot::plot_grid(plotlist = plotList, labels = "AUTO", ncol = 2) +
      patchwork::plot_annotation(
        title = paste("Cluster", target_cluster, "Enrichr Analysis"),
        theme = theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5))
      )
    print(combined_plot)
  }
}

dev.off()
```

Key Findings:

Cluster 1 vs 3. Although both clusters are CD4+ T-cells and look similar on the surface, they are separated by their activation level and biological "mission": cluster 1 cells are in a state of acute inflammation. A very strong TNF-alpha signaling score (over 20 on the -log_10P scale). Cluster 3 cells are also in the tumor area, but they are receiving "calming" or regulatory signals: the TNF-alpha signal is weaker (around 15), and unique pathways like IL-6/STAT3 and IL-2/STAT5 appear.

Cluster 6 (Ambiguous B cells): Likely some B-T cell doublets. It lacks a strong B-cell metabolic signature and sits "physically" between B and T islands on the UMAP, suggesting a mixed transcriptional signal.

Cluster 10 (Plasma B cells): A distinct functional lineage. The extreme shift on the UMAP is driven by a metabolic explosion: massive Proteasome activity and Oxidative Phosphorylation required for high-intensity antibody production.

Cluster 8 represents a specialized population of Vd2 gd T-cells and memory CD8+ T-cells that exhibit a hybrid NK-like transcriptional profile and a high readiness for rapid anti-tumor response, as evidenced by the strong enrichment of T-cell receptor signaling and TNF-alpha inflammatory pathways.

Summary: The clustering is driven by functional shifts (exhaustion, inflammation, and protein secretion) rather than just cell lineage.

```{r}
cellchat <- createCellChat(object = GetAssayData(melanoma, assay = "RNA", slot = "data"), 
                           meta = melanoma@meta.data, 
                           group.by = "cell_type")

cellchat@DB <- CellChatDB.human

cellchat <- cellchat |> 
  subsetData() |> 
  identifyOverExpressedGenes() |> 
  identifyOverExpressedInteractions() |> 
  computeCommunProb(type = "triMean") |>  
  filterCommunication(min.cells = 10) |> 
  computeCommunProbPathway() |> 
  aggregateNet() |>
  netAnalysis_computeCentrality(slot.name = "netP")

pdf("Signaling_Heatmap.pdf", width = 8, height = 10)
netAnalysis_signalingRole_heatmap(cellchat, 
                                  pattern = "outgoing", 
                                  width = 10, 
                                  height = 18, 
                                  font.size = 8)
netAnalysis_signalingRole_heatmap(cellchat, 
                                  pattern = "incoming", 
                                  width = 10, 
                                  height = 18, 
                                 font.size = 8)
dev.off()

rank.df <- data.frame(score = apply(cellchat@netP$prob, 3, sum)) |> 
  (\(df) df[order(df$score, decreasing = TRUE), , drop = FALSE])()
head(rank.df, 15)

melanoma.rank <- subsetCommunication(cellchat, sources.use = "Melanocytes") |>
  aggregate(prob ~ pathway_name, data = _, sum) |>
  (\(df) df[order(df$prob, decreasing = TRUE), ])()
head(melanoma.rank, 15)

pdf("Signaling_diagrams.pdf", width = 16, height = 8)
imm_paths_all <- intersect(c("MIF", "MHC-I", "ADGRE5"), cellchat@netP$pathways)
netVisual_chord_gene(cellchat, signaling = imm_paths_all, 
                     slot.name = "netP", 
                     thresh = 0.1,
                     lab.cex = 0.7, 
                     small.gap = 2)
imm_paths_mel <- intersect(c("MIF", "APP", "LAMININ", "MK", "SPP1", "MHC-I", "CD99"), cellchat@netP$pathways)
netVisual_chord_gene(cellchat, signaling = imm_paths_mel, 
                     slot.name = "netP", 
                     sources.use = "Melanocytes", 
                     thresh = 0.1, 
                     lab.cex = 0.7)
dev.off()
```

Conclusions of the Melanoma Interactome Analysis
- Melanocytes function as the primary signaling hubs, actively coordinating the functional state and communication patterns within the tumor microenvironment.

- The immune compartment demonstrates high recruitment and intense intercellular crosstalk, with CD8+ T-cells and Vd2 gd T-cells serving as the central receivers of signals from all cell clusters.

- Despite active immune infiltration, the high-scoring MIF pathway indicates a dominant state of immunosuppression, effectively neutralizing the cytotoxic potential of the T-cell response.

- Increased activity in LAMININ, CD99, and SPP1 pathways reflects the tumor’s readiness for extracellular matrix remodeling and tissue invasion.

- Elevated expression of APP and MK supports tumor cell survival and growth signals, while the overall transcriptomic profile suggests cells remain primarily in the G1 phase.

- The MHC-I signaling pathway is present but relatively weak compared to immunosuppressive factors, pointing toward impaired antigen presentation and efficient immune evasion by the tumor.



